---
title: "Project3 – Data Science Skills"
author: "Karim Hammoud"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    incremental: true
    widescreen: true
    smaller: true
  openintro::lab_report: default
---

## Introduction
The goal of this project is to predict the US presendential election.


### What are the Most Valued Data Science Skills?

Collaborating and the tream work is one of the valued data science skills, 


### Communication:

We used Github, OneDrive and Slack and Google meeting for collboration.

## the source of the Data

We obtained the data from https://projects.fivethirtyeight.com/2020-election-forecast/

## Libraries
```{r load-packages, message=FALSE}
library(tidyverse)
```

## Loading the data

The data is saved on Github
```{r echo=TRUE}
url <- "https://raw.githubusercontent.com/akarimhammoud/607-Project-3/master/presidential_national_toplines_2020.csv"
polls <- read.csv(file = url, header = TRUE, sep = ",")
head(polls)
```


## The Structure of the data

```{r}
str(polls)
```

## Data subset

We create new framework

```{r}
nat_polls <- subset(polls, select = c(modeldate, candidate_inc, candidate_chal,
             national_voteshare_inc, national_voteshare_chal))
```

## Data Mutate

we genrate new column for

```{r}

colnames(nat_polls) <- c("date", "candidate_Trump", "candidate_Biden", "Trump_Percentage", "Biden_Percentage")
nat_polls
```

## Trump Poll Percentage

Data Visualization

```{r}
plot(nat_polls$Trump_Percentage, type = "b", frame = FALSE, pch = 20,
     col = "red", xlab = "Days", ylab = "Percentage", 
     main = "Trump Projection from June - September 2020",xlim = c(0,100))
```

## Biden Poll Percentage

Data Visualization

```{r}
plot(nat_polls$Biden_Percentage, type = "b", frame = FALSE, pch = 20, 
     col = "blue", xlab = "Days", ylab = "Percentage", 
     main = "Biden Projection from June - September 2020",xlim = c(0,100))
```

## Both Candidates

Data Visualization

```{r}
plot(nat_polls$Biden_Percentage, type="l", col="blue", lwd=2, xlab="Days",
     ylab="Percentage", xlim = c(0,100), ylim = c(44,56))
lines(nat_polls$Trump_Percentage, col="red", lwd=1)
legend("topleft", legend=c("Trump_Percentage", "Biden_Percentage"),
       col=c("red", "blue"), lty=1:1, cex=0.8)
```

## Conclusion

In the past 100 days, the polls shows simillar projects for both candidates, for Trump its around 45% - 47%, on the other side Biden has between 52% - 53%.
