---
title: "Project 3"
author: "Diego Correa"
date: "9/8/2020"
output:
  pdf_document: default
  html_document: default
---

## Overview

```{r include=FALSE}
# Tidyverse to load ggplot and tidyr to manipulate and visualize data frames
library(tidyverse)
# lubridate to clean up the date columns
library(lubridate)
#  RMySQL to load CSV files into MySQL
library(RMySQL)
# for visualization of the US
library(usmap)
# reshape to manipulate dataframe in order to find covariances
library(reshape2)

```


## Loading data

```{r}
#Presidential National Toplines
url1 <- 'https://projects.fivethirtyeight.com/2020-general-data/presidential_national_toplines_2020.csv'
#Presidential State Toplines
url2 <- 'https://projects.fivethirtyeight.com/2020-general-data/presidential_state_toplines_2020.csv'
#Presidential EV Probabilities
url3 <- 'https://projects.fivethirtyeight.com/2020-general-data/presidential_ev_probabilities_2020.csv'
#Presidential Scenario Analysis
url4 <- 'https://projects.fivethirtyeight.com/2020-general-data/presidential_scenario_analysis_2020.csv'
#Economic Index
url5 <- 'https://projects.fivethirtyeight.com/2020-general-data/economic_index.csv'

#Saving data into variables
dfPNT <- read.csv(file = url1)
dfPST <- read.csv(file = url2)
dfPEP <- read.csv(file = url3)
dfPSA <- read.csv(file = url4)
dfEI <- read.csv(file = url5)
```

## Observing & Cleaning 


```{r}
# we notice that the dates in modelcate and timestamp are stored a different format
# than we want to store them for later anaylsis
str(dfEI)

```

When we look at the dataframes, we see the the modeldate and the timestamp fields are stored as character types. Additionally, the timestamp field contains multplie whitespaces. Before dumping the data into the MySQL database, we need to clean up the fields.
As the same process needs to be done to each data frame, a function was created to 
address preform the necessary cleansing mentioned above. We do this with the help of
the lubridate package.


```{r}
# creating function to go into the modeldate and timestamp columns of each csv and converting
# into a format we can perform analysis on
datetime.this <- function(df){
  
  #using the mdy function of lubridate to format the modeldate column
  
  df$modeldate <- mdy(df$modeldate)
  
  #now we need to change the format of the timestamp column
  #using str_replace_all of tidyr, finding 2 whitespaces and replacing with a single one
  df$timestamp <- str_replace_all(df$timestamp, ' {2}', ' ')
  
  #splitting the column into a new data frame where there is a ":" or " "
  x <- data.frame(str_split(df$timestamp, ':| ', n = 4, simplify = TRUE))
  
  #creating a date column by using the dmy function from lubridate
  x <- x %>%
    mutate(date = dmy(X4))
  
  #creating the timestamp column
  x <- x %>%
    mutate(timestamp = as_datetime(paste(date, as.integer(X1), as.integer(X2), as.integer(X3),
                                         sep = ' ')))
  #last step, saving the creating column in the data frame that is passed
  df$timestamp <- x$timestamp
  
  return(df)
}


dfEI <- datetime.this(dfEI)
dfPST <- datetime.this(dfPST)
dfPNT <- datetime.this(dfPNT)
dfPEP <- datetime.this(dfPEP)
dfPSA <- datetime.this(dfPSA)

str(dfEI)

```

## Connecting and storing to db

```{r message=FALSE, include=FALSE}
#storing the data in google cloud database
con <- dbConnect(MySQL(),
                 user = 'root',
                 host = '35.225.135.85',
                 dbname = 'DATA607')

dbWriteTable(con, 'Presidential_National_Toplines', dfPNT, row.names = FALSE, overwrite = TRUE)
dbWriteTable(con, 'Presidential_State_Toplines', dfPST, row.names = FALSE, overwrite = TRUE)
dbWriteTable(con, 'Presidential_EV_Probabilities', dfPEP, row.names = FALSE, overwrite = TRUE)
dbWriteTable(con, 'Presidential_Scenario_Analysis', dfPSA, row.names = FALSE, overwrite = TRUE)
dbWriteTable(con, 'Economic_Index', dfEI, row.names = FALSE, overwrite = TRUE)
```

### Economic Index
The Economic Index table contains economic indicators that serve as inputs to the forecast. 
For more information on these indicators, see this [post](https://fivethirtyeight.com/features/measuring-the-effect-of-the-economy-on-elections/). The economic indexes were collected from the [Federal Reserve Bank Of St. Louis](https://fred.stlouisfed.org/series/DSPIC96) and the stock prices data from [Yahoo Finance](https://finance.yahoo.com/). This sheet contains the following additional columns:


```{r}
#getting the Economic_Index table from MySQL
res_EI <- dbGetQuery(con, 'select modeldate, category, current_zscore, projected_zscore
                  from Economic_Index;')

#modeldate column is save as a character, we need to change the modeldate column in date
res_EI$modeldate <- ymd(res_EI$modeldate)

#graphing the categories in the Economic Index table with a linear regression(geom_smooth)
current <- ggplot(data = res_EI) + 
  geom_smooth(mapping = aes(x = modeldate, y = current_zscore), color = 'brown') +
  geom_smooth(mapping = aes(x = modeldate, y = projected_zscore), color = 'green') +
  facet_wrap(~category) +
  theme(axis.text.x = element_text(angle = 70, hjust = 1))
current + scale_x_date(date_labels = '%m-%Y') +
  labs(title = 'Current & Projected Zscore of Economic Index by Model Date') 
```

## Presidential EV Probabilities table

The Presidential EV Porbabilities table contains the forecasted chances of every possible Electoral College outcome. This table contains only one most recent day's electoral college simulations.

```{r}
str(dfPEP)
```
## Presidential Scenario Analysis

The Presidential Scenario Analysis contains the forecasted chances of various possible election outcome scenarios.


```{r}
str(dfPSA)
```
```{r}
res <- dbGetQuery(con, 'select modeldate, scenario_description, probability from 
                  Presidential_Scenario_Analysis;')

res$modeldate <- ymd(res$modeldate)

probScenario <- ggplot(data = res) +
  geom_line(mapping = aes(modeldate, y = probability, color = scenario_description)) +
  theme(axis.text.x = element_text(angle = 70, hjust = 1))

probScenario + scale_x_date(date_labels = '%m-%Y') + scale_color_discrete(labels = function(x) str_wrap(x, width = 65))
#not very pretty to look at
```
## State Toplines

Looking at red and blue states

```{r}
# grabbing the average probability of the incumbent winning the state
res <- dbGetQuery(con, 'select state, avg(winstate_inc) Inc from Presidential_State_Toplines group by state;')

dfStateWin <- res


plot_usmap(data = dfStateWin, values = 'Inc') + 
  scale_fill_gradient2(
    low = 'blue', high = 'red', mid = 'white', midpoint = .5, limit = c(0,1),
    name = 'Projected Vote Outcomes'
  ) + theme(legend.position = 'right') +
  labs(title = 'Projected Red and Blue States')
```

#### Voter Turnout by State

separating red and blue states based on average probability of winning state

```{r warning=FALSE}
# grabbing the entire presidential state toplines table
res_state <- dbGetQuery(con, 'select * from Presidential_State_Toplines;')

# getting the average probability of the incumbent winning the state
# then filtering and those states that are equal to or greater than 0.50
red_states <- res_state %>%
  group_by(state) %>%
  summarize(avg_winstate_inc = round(mean(winstate_inc, na.rm = TRUE), 3)) %>%
  filter(avg_winstate_inc >= 0.5) %>%
  select(state)

# once I have the red states, I can go back into the original data frame and grab the 
# average projected turnout of those states
red_state_turnout <- res_state %>%
  semi_join(red_states) %>%
  group_by(state) %>%
  summarize(avg_state_turnout = mean(state_turnout, na.rm = TRUE))

# since I have the red states, I can use the anti_join function and grab the average
# turnout of the states that are not red (blue).
blue_state_turnout <- res_state %>%
  anti_join(red_states) %>%
  group_by(state) %>%
  summarize(avg_state_turnout = mean(state_turnout, na.rm = TRUE))

# I add a column in the red & blue state data frame with the value "red" & "blue"
red_state_turnout['state_color'] <- 'red'
blue_state_turnout['state_color'] <- 'blue'

# combine the data frames
df1 <- rbind(red_state_turnout, blue_state_turnout)

# boxplot of the red and blue states' turnout
bp <- ggplot(data = df1, aes(x = state_color, y = avg_state_turnout, fill = state_color)) +
  geom_boxplot() +
  labs(title = 'Projected Turnout of Red & Blue States', subtitle = 'Boxplot')
bp + scale_fill_manual(values = c('blue', 'red'))

# density chart
den <- ggplot(data = df1, aes(x = avg_state_turnout, fill = state_color)) +
  geom_density(alpha = 0.4)
den + scale_fill_manual(values = c('blue', 'red')) +
  labs(title = 'Projected Turnout of Red & Blue States', subtitle = 'Density')
```
## States with the Most Influence

To find the answer to this, I added two columns found in the Presidential_State_Toplines table,
"win_EC_if_win_state_inc" & "win_EC_if_win_state_chal".

```{r}
# creating a column that is the sum of the two columns, group them, find the average, and order them
states_influence <- res_state %>%
  mutate(total_EC_influence = win_EC_if_win_state_inc + win_EC_if_win_state_chal) %>% 
  group_by(state) %>%
  summarize(avg_EC_Influence = mean(total_EC_influence, na.rm = TRUE)) %>%
  arrange(desc(avg_EC_Influence))

# plotting on a map
plot_usmap(data = states_influence, values = 'avg_EC_Influence') + 
  scale_fill_continuous(
    low = 'white', high = 'purple', name = 'Wins EC if Wins State Prob'
  ) + theme(legend.position = 'right') +
  labs(title = 'Wins EC if Wins State Prob')

# looking at the dataframe
head(states_influence)


```
## Economic Index & Presidential National Toplines

Looking at the economic index and the probabilities of winning the presidency

```{r}
# joining the tables on the modeldate
res_Nat_EI <- dbGetQuery(con, 'select p.modeldate, ecwin_inc, ecwin_chal, 
                            category, indicator, projected_zscore, current_zscore
                             from 
                             Presidential_National_Toplines p 
                             inner join  Economic_Index ei
                                  on ei.modeldate = p.modeldate;')

# plotting the relationship of the EI's and the prob of the incumbent winning
ggplot(data = res_Nat_EI, aes(x = ecwin_inc, y = current_zscore, color = 'red')) +
  geom_smooth() + 
  facet_wrap(~ category) +
  theme(legend.position =  'none', axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(title = 'Probability of Incumbent Winning vs Current Zscore of Economic Index')

# plotting the relationship of the EI's and the prob of the challenger winning
ggplot(data = res_Nat_EI, aes(x = ecwin_chal, y = current_zscore)) +
  geom_smooth(aes(fill = 'blue')) + 
  facet_wrap(~ category) +
  theme(legend.position =  'none', axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(title = 'Probability of Challenger Winning vs Current Zscore of Economic Index')

# we see that income is positively correlated with the incumbent winning and 
# spending positively correlated with the challenger winning
```

## Covariances of variables in the Presidential National Toplines



```{r}
# grabbing entire presidential national toplines, has 40 columns
res_nat <- dbGetQuery(con, 'select * from Presidential_National_Toplines;')

# filter out the columns that we need, only the columns ending "chal" and "inc"
# first finding the column names with "chal" for challenger
col  <- colnames(res_nat)
chal_col <- str_match(col, '.*chal$')
chal_col <- chal_col[!is.na(chal_col)]

# saving the "chal" columns in a data frame
dfChal <- res_nat[, chal_col[2:length(chal_col)]]

# finding the correlations
chal_cormatrix <- round(cor(dfChal), 2)

# finding the column names with "inc" for incumbent
inc_col <- str_match(col, '.*inc$')
inc_col <- inc_col[!is.na(inc_col)]

# saving the "inc" columns in a data frame
dfInc <- res_nat[, inc_col[2:length(inc_col)]]

# finding the correlations
inc_cormatrix <- round(cor(dfInc), 2)


# function to remove the bottom right triangle of matrix as it is redundant
get_upper_tri <- function(cormatrix){
  cormatrix[lower.tri(cormatrix)] <- NA
  return(cormatrix)
}

# function to reorder the matrix from hi to low covariance
reorder_cormatrix <- function(cormatrix){
  dd <- as.dist((1 - cormatrix) / 2)
  hc <- hclust(dd)
  cormatrix <- cormatrix[hc$order, hc$order]
}

# running the data frames through the functions
chal_cormatrix <- reorder_cormatrix(chal_cormatrix)
chal_upper_triangle <- get_upper_tri(chal_cormatrix)
inc_cormatrix <- reorder_cormatrix(inc_cormatrix)
inc_upper_triangle <- get_upper_tri(inc_cormatrix)

# the melt function from reshape2 library reshapes the matrix for it to be plotted
chal_melted_cormatrix <- melt(chal_upper_triangle, na.rm = TRUE)
inc_melted_cormatrix <- melt(inc_upper_triangle, na.rm = TRUE)

# plotting the challenger variables
ggplot(data = chal_melted_cormatrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = 'white') +
  scale_fill_gradient2(low = 'white', high = 'red', 
                       midpoint = 0.5, limit = c(0,1), space = 'Lab') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  coord_fixed() +
  labs(title = 'Covariance of Challenger Measures')

# plotting the incumbent variables
ggplot(data = inc_melted_cormatrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = 'white') +
  scale_fill_gradient2(low = 'white', high = 'red', 
                       midpoint = 0.5, limit = c(0,1), space = 'Lab') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  coord_fixed() + 
  labs(title = 'Covariance of Incumbent Measures')

# As you guessed it, the popwin variable for both the challenger and incumbent has the 
# lowest covariance

```
## Presidential National Toplines

contains the final national topline on each day.Chances to win electoral votes (Biden in blue, Trump in red)


```{r}
# grabbing the probabilities of the incumbent and challenger winning the presidency
res_nat_EC <- dbGetQuery(con, 'select modeldate, ecwin_inc, ecwin_chal from 
                  Presidential_National_Toplines;')

# changing the modeldate from character to date using lubridate
res_nat_EC$modeldate <- ymd(res_nat_EC$modeldate)

# graphing the two columns by dates
probEWin <- ggplot(data = res_nat_EC) +
  geom_line(mapping = aes(x = modeldate, y = ecwin_inc), color = 'red') +
  geom_line(mapping = aes(x = modeldate, y = ecwin_chal), color = 'blue') +
  theme(axis.text.x =  element_text(angle = 70, hjust = 1))

probEWin + scale_x_date(date_labels = '%m-%Y') +
  labs(title = 'Probability of Incumbent & Challenger Winning Election' , y = 'probability')


```



